Array

### 2014-02-27 DOCUMENTATION/AdvancedTypes.tex
Adding structure support details.

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\ensuremath{\downarrow}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\developed}[0]{\ensuremath{\textsf{developed}}}
\newcommand{\stateless}[0]{\ensuremath{\textsf{stateless}}}
\newcommand{\importedmodules}[0]{\ensuremath{\textsf{i-mod}}}
\newcommand{\trimportedmodules}[0]{\ensuremath{\textsf{i-mod$\mathsf{\ast}$}}}
\newcommand{\importedinstances}[0]{\ensuremath{\textsf{i-inst}}}
\newcommand{\trimportedinstances}[0]{\ensuremath{\textsf{i-inst$\mathsf{\ast}$}}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

@@

\newcommand{\stateindex}[0]{\ensuremath{\sf\tau}}

@@

\begin{figure}[H]
  \begin{center}
    {\footnotesize
      \frame{
    \begin{tabular}[t]{ccc}
      \begin{tabular}[t]{rcl}
        \hline
        \multicolumn{3}{|c|}{Machine: \B{Machine}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{const} & : & seq \B{Cons} \\
        \B{var} & : & seq \B{Vari} \\
        \B{init} & : & seq \B{Inst} \\
        \B{op} & : & seq \B{Oper} \\
        \B{impl} & : & opt \B{Impl} \\
        \hline
        \multicolumn{3}{|c|}{Implementation: \B{Impl}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{impo} & : & seq \B{Import} \\
        \B{const} & : & seq \B{Cons} \\
        \B{var} & : & seq \B{Vari} \\
        \B{init} & : & seq \B{Inst} \\
        \B{op} & : & seq \B{Oper} \\
        \hline
        \multicolumn{3}{|c|}{Imports: \B{Import}} \\
        \hline
        \B{mach} & : & \B{Machine} \\
        \B{pre} & : & opt \B{Name}
        \\
        \hline
        \multicolumn{3}{|c|}{Constant: \B{Cons}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{type} & : & \B{Type} \\
        \B{val} & : & \B{Value} \\
        \hline
        \multicolumn{3}{|c|}{Variable: \B{Vari}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{type} & : & \B{Type} \\
        \B{scope} & : & \B{Impl} \alt \B{Oper}\\
	\hline
        \multicolumn{3}{|c|}{Operation: \B{Oper}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{inp} & : & seq \B{Vari} \\
        \B{out} & : & seq \B{Vari} \\
        \B{body} & : & \B{Inst}
        \\
       \hline
        \multicolumn{3}{|c|}{Instruction: \B{Inst}} \\
        \hline
        \multicolumn{3}{c}{\B{Blk} \alt \B{VarD} \alt \B{If} \alt \B{BEq} \alt } \\
        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case} \alt \B{Skip}} \\
        \hline
        \multicolumn{3}{|c|}{Block: \B{Blk}} \\
        \hline
        \B{body} & : & seq \B{Inst} \\
	\end{tabular}

       \begin{tabular}[t]{rcl}   
        \hline
        \multicolumn{3}{|c|}{Variable declaration: \B{VarD}} \\
        \hline
        \B{vars} & : & seq \B{Name} \\
        \B{body} & : & seq \B{Inst} \\
  	\hline
        \multicolumn{3}{|c|}{If: \B{If}} \\
        \hline
        \B{branches} & : & seq \B{IfBr} \\
        \hline
        \multicolumn{3}{|c|}{Becomes equal: \B{Beq}} \\
        \hline
        \B{lhs} & : & \B{LValue} \\
        \B{rhs} & : & \B{Expr} \\
	\hline
	\multicolumn{3}{|c|}{Label value: \B{LValue}} \\	
	\hline
	\multicolumn{3}{c}{ \B{Vari} \alt \B{ArrayItem} }\\
        \hline
        \multicolumn{3}{|c|}{Operation call: \B{Call}} \\
        \hline
        \B{op} & : & \B{Oper} \\
        \B{inp} & : & seq \B{Expr} \\
        \B{out} & : & seq \B{Name} \\
        \B{inst} & : & opt \B{Impo} \\
        \hline
        \multicolumn{3}{|c|}{Case conditional: \B{Case}} \\
        \hline
        \B{expr} & : & \B{Expr} \\
        \B{branches} & : & seq \B{CaseBr}\\
	\hline
        \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
        \hline
        \B{val} & : & seq \B{Value} \\
        \B{body} & : & \B{Inst}
        \\
        \hline
        \multicolumn{3}{|c|}{While loop: \B{While}} \\
        \hline
        \B{cond} & : & \B{Pred} \\
        \B{body} & : & \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{If branch: \B{IfBr}} \\
        \hline
        \B{cond} & : & opt \B{Pred} \\
        \B{body} & : & \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{Expression: \B{Expr}} \\
        \hline
        \multicolumn{3}{c} {\B{Lit} \alt \B{Term} \alt \B{Pred} \alt \B{Cons}}\\
	\multicolumn{3}{c} {\B{Vari} \alt \B{Array} \alt \B{ArrayItem} } \\
        \hline
        \multicolumn{3}{|c|}{Term expression: \B{Term}} \\
        \hline
        \B{op} & : & \B{ArithOp} \\
        \B{args} & : & seq \B{Exp} \\
	\hline
        \multicolumn{3}{|c|}{Predicate: \B{Pred}} \\
        \hline
        \multicolumn{3}{c}{\B{Form} \alt \B{Comp}} \\
\end{tabular}

  \begin{tabular}[t]{rcl}
	\hline
        \multicolumn{3}{|c|}{Boolean formula: \B{Form}} \\
        \hline
        \B{op} & : & \B{BoolOp} \\
        \B{args} & : & seq \B{Pred} \\
        \hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{Struct} \alt \B{ArrayType}}\\
	\hline
  	\multicolumn{3}{|c|}{Data structure : \B{Struct}} \\
        \hline
	\B{id} & : &  \B{Name}\\
        \B{elements} & : & seq \B{Name} \\
	\B{type} & : & seq \B{Type}\\
	\hline
%  	\multicolumn{3}{|c|}{Data record : \B{Record}} \\
%        \hline
%	\B{id} & : &  \B{Name}\\
%        \B{elements} & : & seq \B{Name} \\
%	\B{type} & : & \B{Struct}\\
%	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Expr}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Expr} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).
      Sequence and optional attributes are denoted ``seq'' and ``opt'',
      respectively}
    \label{tab:node-attr}
  \end{center}
\end{figure}

@@

Two approaches are possible to support arrays. An approach is importing
reusable libraries of AtelierB that are reported in the official document of
ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}.
This approach is supported, but the generated code is more complex and it needs
more intermediate steps to manipulate the data of imported components. Moreover,
this approach allows dynamic allocation for arrays, but it is strongly not
recommended to safety-critical systems. A second approach enjoys the direct 
correspondence between B0 native arrays and LLVM-IR and this approach is presented here.

A concrete array is a total function where the domain indexes the range 
elements. The domain is composed of Cartesian product of sequence of concrete 
sets. The node Data array in AST can represent a declaration of array or a 
selection of element. The first element ``dec'' declares explicitly an array or 
references a variable array. The second element ``type'' defines a sequence of 
array types, where the last element defines the range of function and the other 
elements define the domain. The third element ``indexes'' is optional, and it 
represents the selection of an element in array, or the same type structure 
when it is a structure.

@@

The essential elements of the array are its name, size and type. 
The size of ``e'' array with the indexes (i1, ... , in) are defined according 
to the formula: $size(e) = (e.i1\_end - e.i1\_star+1) * ... * (e.in\_end - 
e.in\_start+1)$.

@@

A list of type and variable is generated by the following rule.
%TODO: insert a rule to do the shift address with the first element of interval
\begin{align*}
&\trad{n}{LRExp} \isdef \LET p \sep v \sep t  = \trad{n}{ShiftExp}  \IN  \llvm{",\PH{t} \PH{v} "} \sep p\\
&\trad{n l}{LRExp} \isdef \LET r_1 \sep p_1 = \trad{n}{LRExp} \AND 
 \LET r_2 \sep p_2 = \trad{l}{LRExp} \IN\\
& \quad  \llvm{"\PH{r_1} \PH{r_2}"} \sep \llvm{"\PH{p_1} \PH{p_2}"}\\
\end{align*}

@@

\paragraph{Restriction on array and structure:} The translation of direct attribution
 between array expression is not yet supported, for example:

\begin{pascalcode}
 array1 := array2;
 structure1 := structure2;
\end{pascalcode}

The B0 language supports attribution directly between arrays or structures, even though the attribution
must have the same domain type and range type.

@@

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}
MACHINE Rec
SETS
 ID={aa,bb}
VARIABLES xx, bank, total
INVARIANT
 xx : struct(name : ID, balance : NAT) & bank : struct(name : ID) &
 total:struct(money:NAT)
INITIALISATION 
 bank := rec(name:bb) ||
 xx:= rec(name:aa, balance:10) ||
 total := rec(money:20)
OPERATIONS

@@

  unsafe_dec = 
  BEGIN
       xx'balance := xx'balance - 1
  END
END
\end{minted}

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}
IMPLEMENTATION
    Rec_i

@@

  unsafe_dec = BEGIN
       xx'balance := xx'balance -1 
  END
END
\end{minted}

@@

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}

MACHINE
   Array
VARIABLES arr

@@

END
\end{minted}

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{c}
IMPLEMENTATION 
   Array_i
REFINES
   Array
CONCRETE_VARIABLES
   arr
INVARIANT 
    arr :  ((0..99) --> 0..1000)
INITIALISATION
   arr(1) := 100
OPERATIONS
   set ( ix , tt ) =
   IF
       ix >= 0 & ix <=99 & tt >= 0  & tt <= 1000
   THEN

@@

       VAR temp IN
           temp := arr(jx);
           arr(jx) := arr(ix);
           arr(ix) := temp
       END
   END

\end{minted}

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{llvm}

@Array__arr = internal global [100 x i32] zeroinitializer, align 16

@@

; <label>:14                                      ; preds = %11
  %15 = load i32* %2, align 4
  %16 = load i32* %1, align 4
  %17 = sext i32 %16 to i64
  %18 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %17
  store i32 %15, i32* %18
  br label %19
; <label>:19                                      ; preds = %14, %11, %8, %5, %0
  ret void
}
\end{minted}

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{llvm}

define void @Array__read(i32 %ix, i32* %tt) nounwind ssp {
  %1 = alloca i32, align 4
  %2 = alloca i32*, align 8
  store i32 %ix, i32* %1, align 4
  store i32* %tt, i32** %2, align 8
  %3 = load i32* %1, align 4
  %4 = icmp sge i32 %3, 0
  br i1 %4, label %5, label %14

@@

; <label>:16                                      ; preds = %14, %8
  ret void
}
\end{minted}

\begin{minted}[frame=single,bgcolor=bg,linenos=true,fontsize=\scriptsize]{llvm}

define void @Array__swap(i32 %ix, i32 %jx) nounwind ssp {
  %1 = alloca i32, align 4
  %2 = alloca i32, align 4
  %temp = alloca i32, align 4
  store i32 %ix, i32* %1, align 4
  store i32 %jx, i32* %2, align 4
  %3 = load i32* %1, align 4
  %4 = icmp sge i32 %3, 0
  br i1 %4, label %5, label %30

@@

\end{minted}

@@

	INVARIANT 1>0
	VARIANT nn-ii
	END
   END
END

\end{minted}

>>>>

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\ensuremath{\downarrow}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\developed}[0]{\ensuremath{\textsf{developed}}}
\newcommand{\stateless}[0]{\ensuremath{\textsf{stateless}}}
\newcommand{\importedmodules}[0]{\ensuremath{\textsf{i-mod}}}
\newcommand{\trimportedmodules}[0]{\ensuremath{\textsf{i-mod$\mathsf{\ast}$}}}
\newcommand{\importedinstances}[0]{\ensuremath{\textsf{i-inst}}}
\newcommand{\trimportedinstances}[0]{\ensuremath{\textsf{i-inst$\mathsf{\ast}$}}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}
\newcommand{\SBPos}[0]{\ensuremath{\sf\varphi}}
\newcommand{\SBType}[0]{\ensuremath{\sf\tau}}

@@

@@

\begin{figure}[H]
  \begin{center}
    {\footnotesize
      \frame{
    \begin{tabular}[t]{ccc}
      \begin{tabular}[t]{rcl}
        \hline
        \multicolumn{3}{|c|}{Machine: \B{Machine}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{const} & : & seq \B{Cons} \\
        \B{var} & : & seq \B{Vari} \\
        \B{init} & : & seq \B{Inst} \\
        \B{op} & : & seq \B{Oper} \\
        \B{impl} & : & opt \B{Impl} \\
        \hline
        \multicolumn{3}{|c|}{Implementation: \B{Impl}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{impo} & : & seq \B{Import} \\
        \B{const} & : & seq \B{Cons} \\
        \B{var} & : & seq \B{Vari} \\
        \B{init} & : & seq \B{Inst} \\
        \B{op} & : & seq \B{Oper} \\
        \hline
        \multicolumn{3}{|c|}{Imports: \B{Import}} \\
        \hline
        \B{mach} & : & \B{Machine} \\
        \B{pre} & : & opt \B{Name}
        \\
        \hline
        \multicolumn{3}{|c|}{Constant: \B{Cons}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{type} & : & \B{Type} \\
        \B{val} & : & \B{Value} \\
        \hline
        \multicolumn{3}{|c|}{Variable: \B{Vari}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{type} & : & \B{Type} \\
        \B{scope} & : & \B{Impl} \alt \B{Oper}\\
	\hline
        \multicolumn{3}{|c|}{Operation: \B{Oper}} \\
        \hline
        \B{id} & : & \B{Name} \\
        \B{inp} & : & seq \B{Vari} \\
        \B{out} & : & seq \B{Vari} \\
        \B{body} & : & \B{Inst}
        \\
       \hline
        \multicolumn{3}{|c|}{Instruction: \B{Inst}} \\
        \hline
        \multicolumn{3}{c}{\B{Blk} \alt \B{VarD} \alt \B{If} \alt \B{BEq} \alt } \\
        \multicolumn{3}{c}{\B{Call} \alt \B{While} \alt \B{Case} \alt \B{Skip}} \\
        \hline
        \multicolumn{3}{|c|}{Block: \B{Blk}} \\
        \hline
        \B{body} & : & seq \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{Variable declaration: \B{VarD}} \\
        \hline
        \B{vars} & : & seq \B{Name} \\
        \B{body} & : & seq \B{Inst} \\
\end{tabular}
\begin{tabular}[t]{rcl}   
  	\hline
        \multicolumn{3}{|c|}{If: \B{If}} \\
        \hline
        \B{branches} & : & seq \B{IfBr} \\
        \hline
        \multicolumn{3}{|c|}{Becomes equal: \B{Beq}} \\
        \hline
        \B{lhs} & : & \B{LValue} \\
        \B{rhs} & : & \B{Expr} \\
	\hline
	\multicolumn{3}{|c|}{Label value: \B{LValue}} \\	
	\hline
	\multicolumn{3}{c}{ \B{Vari} \alt \B{ArrayItem} }\\
        \hline
        \multicolumn{3}{|c|}{Operation call: \B{Call}} \\
        \hline
        \B{op} & : & \B{Oper} \\
        \B{inp} & : & seq \B{Expr} \\
        \B{out} & : & seq \B{Name} \\
        \B{inst} & : & opt \B{Impo} \\
        \hline
        \multicolumn{3}{|c|}{Case conditional: \B{Case}} \\
        \hline
        \B{expr} & : & \B{Expr} \\
        \B{branches} & : & seq \B{CaseBr}\\
	\hline
        \multicolumn{3}{|c|}{Case branch: \B{CaseBr}} \\
        \hline
        \B{val} & : & seq \B{Value} \\
        \B{body} & : & \B{Inst}
        \\
        \hline
        \multicolumn{3}{|c|}{While loop: \B{While}} \\
        \hline
        \B{cond} & : & \B{Pred} \\
        \B{body} & : & \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{If branch: \B{IfBr}} \\
        \hline
        \B{cond} & : & opt \B{Pred} \\
        \B{body} & : & \B{Inst} \\
        \hline
        \multicolumn{3}{|c|}{Expression: \B{Expr}} \\
        \hline
        \multicolumn{3}{c} {\B{Lit} \alt \B{Term} \alt \B{Pred} \alt \B{Cons}}\\
	\multicolumn{3}{c} {\B{Vari} \alt \B{Array} \alt \B{ArrayItem} } \\
        \hline
        \multicolumn{3}{|c|}{Term expression: \B{Term}} \\
        \hline
        \B{op} & : & \B{ArithOp} \\
        \B{args} & : & seq \B{Exp} \\
	\hline
        \multicolumn{3}{|c|}{Predicate: \B{Pred}} \\
        \hline
        \multicolumn{3}{c}{\B{Form} \alt \B{Comp}} \\
	\hline
        \multicolumn{3}{|c|}{Boolean formula: \B{Form}} \\
        \hline
        \B{op} & : & \B{BoolOp} \\
        \B{args} & : & seq \B{Pred} \\
\end{tabular}

\begin{tabular}[t]{rcl}
        \hline
        \multicolumn{3}{|c|}{Comparison: \B{Comp}} \\
        \hline
        \B{op} & : & \B{RelOp} \\
        \B{arg1}, \B{arg2} & : & \B{Exp} \\
        \hline
        \multicolumn{3}{|c|}{Data type: \B{Type}} \\
        \hline
        \multicolumn{3}{c}{\B{INT} \alt \B{BOOL}\alt}\\
	\multicolumn{3}{c}{\B{Struct} \alt \B{ArrayType}}\\
	\hline
  	\multicolumn{3}{|c|}{Data structure : \B{Struct}} \\
        \hline
	\B{id} & : &  \B{Name}\\
        \B{elements} & : & seq \B{Name} \\
	\B{type} & : & seq \B{Type}\\
	\hline
	\multicolumn{3}{|c|}{Access struct : \B{StructItem}} \\
        \hline
	\B{base} & : &  \B{Struct} \\
	\B{element} & : & \B{Name}\\
	\B{type} & : & \B{Type}\\
	\hline
        \multicolumn{3}{|c|}{Array type : \B{ArrayType}} \\
        \hline 
	\B{dom} & : & seq \B{Set}\\
	\B{ran} & : & \B{Set}\\
        \hline
        \multicolumn{3}{|c|}{Array expression : \B{Array}} \\
        \hline
	\multicolumn{3}{c}{ seq \B{Map} \alt \B{Product} \alt \B{Name}  }\\
	\hline
  	\multicolumn{3}{|c|}{Ordered pair : \B{Map}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 
	\hline
  	\multicolumn{3}{|c|}{Cartesian product : \B{Product}} \\
        \hline
	\B{arg1} & : & seq \B{Expr}\\ 
	\B{arg2} & : & \B{Expr}\\ 	
	\hline
	\multicolumn{3}{|c|}{Access array : \B{ArrayItem}} \\
        \hline
	\B{base} & : &  \B{Array} \\
	\B{indexes} & : & seq \B{Expr}\\
	\B{type} & : & \B{ArrayType}\\
	\hline
	\multicolumn{3}{|c|}{Simple set : \B{Set}} \\
	\hline
	\multicolumn{3}{c}{\B{BOOL} \alt \B{NAT} \alt \B{NAT$_{1}$}\alt } \\
	\multicolumn{3}{c}{\B{INT} \alt \B{Interval}} \\
	\hline
	\multicolumn{3}{|c|}{B0 interval : \B{Interval}} \\
	\hline
	\B{start}, \B{end} & : & \B{Expr} \\
	\hline
	\multicolumn{3}{|c|}{Literal : \B{Lit}} \\
	\hline
	\multicolumn{3}{c}{ \B{Bool\_lit} \alt \B{Int\_lit}} \\ 
      \end{tabular}
    \end{tabular}
  }
  }
    \caption{Abstract syntax structure of a B implementation.  For each abstract
      syntax element, we give the name of the class (e.g. \B{CaseBr} for a
      branch in a case instruction, and the different attributes of the class),
      or a list of the possible sub-classes (e.g. \B{Inst} for the different
      kinds of instructions).
      Sequence and optional attributes are denoted ``seq'' and ``opt'',
      respectively}
    \label{tab:node-attr}
  \end{center}
\end{figure}

@@

Two approaches are possible to support arrays and structures. An approach is importing reusable libraries of AtelierB that are reported in the official document of ClearSy\footnote{\url{http://www.atelierb.eu/ressources/DOC/english/reusable-components-reference.pdf}}. This approach is supported, but the generated code is more complex and it needs more intermediate steps to manipulate the data of imported components. Moreover, this approach allows dynamic allocation for arrays, but it is strongly not recommended to safety-critical systems. A second approach enjoys the direct correspondence between B0 native arrays and LLVM-IR and this approach is presented here.

\section{Structure}
The struct declaration defines a list of variables to be placed under one name, 
allowing several variables to be accessed by one base name. This example get andset one value of xx'balance and initializes a structure.

\begin{ccode}
xx := rec(name:aa, balance:10) ;
xx'balance := xx'balance -1;
\end{ccode}

The following rule get an element value from the struct. The same rule is used in left side of attribution and right side of attribution. The functions \SBPos(n) and \SBType(n)  get respectively the position and type of $n$ element in the structure. 
\begin{align*}
\begin{split}
  \trad{n}{\B{StructItem}} \isdef \\
   & \LET \ell = \lbl \AND \PH{t} = n.type \AND \PH{v} = n.element \IN\\
   & \llvm{"\PH{\ell} = getelementptr \%\PH{t} @\PH{v}, i32 0, i32  \SBPos(\B{n})"} \\
   & \sep  \ell  \sep \SBType(\B{n})  \\
\end{split}
\end{align*}

AST

record structure : Record
elements : seq Name;
values : seq Expr;

Isto é uma expressão

To remove the restrictions:
( ) The direct attribution between arrays  is madded in similar way

\section{Arrays}
%TODO: update according the new names in AST, just change the names.
A concrete array is a total function where the domain indexes the range 
elements. The domain is composed of Cartesian product of sequence of concrete 
sets. The node Array in AST represents a declaration of array or a 
selection of element. The first element ``dec'' declares explicitly an array or 
references a variable array. The second element ``type'' defines a sequence of 
array types, where the last element defines the range of function and the other 
elements define the domain. The third element ``indexes'' is optional, and it 
represents the selection of an element in array, or the same type structure 
when it is a structure.

@@

The essential elements of the array are its name, size and type. 
The size of ``e'' array with the indexes (i1,...,in) are defined according 
to the formula: $size(e) = (e.i1\_end - e.i1\_star+1) * ... * (e.in\_end - 
e.in\_start+1)$.

@@

A list of type and variable is generated by the following rule.
\begin{align*}
&\trad{n}{LRExp} \isdef \LET p \sep v \sep t  = \trad{n}{ShiftExp}  \IN  \llvm{",\PH{t} \PH{v} "} \sep p\\
&\trad{n l}{LRExp} \isdef \LET r_1 \sep p_1 = \trad{n}{LRExp} \AND 
 \LET r_2 \sep p_2 = \trad{l}{LRExp} \IN\\
& \quad  \llvm{"\PH{r_1} \PH{r_2}"} \sep \llvm{"\PH{p_1} \PH{p_2}"}\\
\end{align*}

@@

\paragraph{Restriction on array:} The translation of direct attribution
 between array expression is not yet supported, for example:
\begin{pascalcode}
 array1 := array2;
\end{pascalcode}

@@

\begin{ccode}
MACHINE Rec
SETS
 ID={aa,bb}
VARIABLES xx, bank, total
INVARIANT
 xx : struct(name : ID, balance : NAT) & bank : struct(name : ID) &
 total:struct(money:NAT)
INITIALISATION 
 bank := rec(name:bb) ||
 xx:= rec(name:aa, balance:10) ||
 total := rec(money:20)
OPERATIONS

@@

  unsafe_dec = 
  BEGIN
       xx'balance := xx'balance - 1
  END
END
\end{ccode}

\begin{ccode}
IMPLEMENTATION
    Rec_i

@@

  unsafe_dec = BEGIN
       xx'balance := xx'balance -1 
  END
END
\end{ccode}

@@

\begin{ccode}
MACHINE
   Array
VARIABLES arr

@@

END
\end{ccode}

\begin{ccode}
IMPLEMENTATION 
   Array_i
REFINES
   Array
CONCRETE_VARIABLES
   arr
INVARIANT 
    arr :  ((0..99) --> 0..1000)
INITIALISATION
   arr(1) := 100
OPERATIONS
   set ( ix , tt ) =
   IF
       ix >= 0 & ix <=99 & tt >= 0  & tt <= 1000
   THEN

@@

       VAR temp IN
           temp := arr(jx);
           arr(jx) := arr(ix);
           arr(ix) := temp
       END
   END
END
\end{ccode}

\begin{ccode}
@Array__arr = internal global [100 x i32] zeroinitializer, align 16

@@

; <label>:14                                      ; preds = %11
  %15 = load i32* %2, align 4
  %16 = load i32* %1, align 4
  %17 = sext i32 %16 to i64
  %18 = getelementptr  [100 x i32]* @Array__arr, i32 0, i64 %17
  store i32 %15, i32* %18
  br label %19
; <label>:19                                      ; preds = %14, %11, %8, %5, %0
  ret void
}
\end{ccode}

\begin{llvmcode}
define void @Array__read(i32 %ix, i32* %tt) nounwind ssp {
  %1 = alloca i32, align 4
  %2 = alloca i32*, align 8
  store i32 %ix, i32* %1, align 4
  store i32* %tt, i32** %2, align 8
  %3 = load i32* %1, align 4
  %4 = icmp sge i32 %3, 0
  br i1 %4, label %5, label %14

@@

; <label>:16                                      ; preds = %14, %8
  ret void
}
\end{llvmcode}

\begin{llvmcode}
define void @Array__swap(i32 %ix, i32 %jx) nounwind ssp {
  %1 = alloca i32, align 4
  %2 = alloca i32, align 4
  %temp = alloca i32, align 4
  store i32 %ix, i32* %1, align 4
  store i32 %jx, i32* %2, align 4
  %3 = load i32* %1, align 4
  %4 = icmp sge i32 %3, 0
  br i1 %4, label %5, label %30

@@

\end{llvmcode}

@@

	INVARIANT 1>0
	VARIANT nn-ii
	END
   END
END
\end{ccode}
